Food-web complexity flattens the fitness landscape of an insect herbivore

### 2018-06-29 elsevier_test/elsevier_test.Rmd
adding code directly to file.

@@

To characterize the shape of the fitness landscape, we quantified selection gradients acting on each trait in simple vs. complex food webs. We did this by fitting separate statistical models to data from each food-web treatment. We used generalized linear mixed models (GLMMs, @Bolker2009) with larval survival (0 or 1) as our response variable and measure of fitness. We specified linear and quadratic terms for each gall trait as well as linear interaction terms between each gall trait, as fixed effects in the statistical models. To account for the correlated structure of clutch size (gall level) and female preference (plant level) as well as any other independent effects of willow genotype on larval survival, we specified gall ID nested within plant ID nested within plant genotype as random intercepts in our statistical models. Since we were interested in characterizing the fitness landscape -- the relationship between mean trait values and population mean fitness -- we assumed the mean value of our random effects (i.e. setting them to zero) to estimate selection gradients. We then used the method of @Janzen1998 to calculate directional ($\beta_{z_i}$), nonlinear ($\gamma_{z_i}$), and correlational ($\gamma_{z_i,z_j}$) selection gradients and used parametric bootstrapping (1000 replicates) to calculate their 95% confidence intervals (@Bolker2009). To test whether selection gradients differed between treatments, we used our bootstrapped estimates to calculate the probability that selection gradients in the simple food web were larger/smaller than in the complex food web (i.e. the p-value).

>>>>

```{r setup, include=FALSE}

## Load required libraries ----
library(tidyverse)
library(cowplot) # pretty default ggplots
library(broom) # for tidying multiple linear models
library(mgcv) # generalized additive models
library(gamm4) # generalized additive mixed-models
library(gsg) # calculating selection gradients
library(visreg)
library(viridis) # for color palette

## Load and manage data ----

gall_selection.df <- read_csv("gall_selection_data.csv") %>%
  # convert appropriate variables to characters instead of integers
  mutate(Plant_Position = as.character(Plant_Position),
         Gall_Number = as.character(Gall_Number)) %>%
  unite(Gall_ID, Gall_Number, Gall_Letter, remove = FALSE) %>%

  # subset data for analysis
  filter(phenology == "early", Location == "tree",
         platy > 0 | ectos > 0 | pupa > 0) %>%                  # eliminate unknown sources of mortality
  mutate(gall_survival = as.numeric(ifelse(pupa > 0, 1, 0)),
         egg_parasitoid = as.numeric(ifelse(platy > 0, 1, 0)))

# convert treatment to factor to use as 'by' variable in generalized additive models
gall_selection.df <- gall_selection.df %>%
  mutate(Treatment_contr.sum=factor(Treatment.focus), 
         sc.gall_size = as.numeric(scale(Gall_Height_mm)),
         sc.clutch_size = as.numeric(scale(gall_individuals)),
         sc.female_preference = as.numeric(scale(Density_per_100_shoots)))
contrasts(gall_selection.df$Treatment_contr.sum) <- 'contr.sum'

## Create separate datasets for quantifying selection gradients ----
control_df <- gall_selection.df %>%
  filter(Treatment.focus == "Control") %>%
  mutate(sc.gall_size = (Gall_Height_mm - mean(Gall_Height_mm))/sd(Gall_Height_mm),
         sc.clutch_size = (gall_individuals - mean(gall_individuals))/sd(gall_individuals),
         sc.female_preference = (Density_per_100_shoots - mean(Density_per_100_shoots))/sd(Density_per_100_shoots))

treatment_df <- gall_selection.df %>%
  filter(Treatment.focus == "Ectoparasitoid exclusion") %>%
  mutate(sc.gall_size = (Gall_Height_mm - mean(Gall_Height_mm))/sd(Gall_Height_mm),
         sc.clutch_size = (gall_individuals - mean(gall_individuals))/sd(gall_individuals),
         sc.female_preference = (Density_per_100_shoots - mean(Density_per_100_shoots))/sd(Density_per_100_shoots))

## Custom functions to avoid repetitive code ----
gamm.model <- function(formula, data) {
  gamm4(formula, data, 
        random = ~(1|Genotype/Plant_Position/Gall_Number), 
        family = binomial(link = logit))
}

gam.plot <- function(gam.model, ...) {
  plot(gam.model, se = TRUE, seWithMean = T, shift = mean(predict(gam.model)), trans = function(x) {exp(x)/(1+exp(x))}, ...)
}

# experimenting with boot.case gave bootstrapped distributions that weren't centered near the model mean estimate.
# also, using 'refit.smooth = T' also generated some very skewed distributions
# the recipe below provided the most sensible estimates for my data
gradient.calc <- function(mod, phenotype, covariates = NULL) {
  gam.gradients(mod, phenotype, covariates, parallel = 'multicore', ncpus = 32, 
                refit.smooth = F,
                n.boot = 1000,
                se.method = 'boot.para', #'posterior', #'boot.para', #'permute', 'boot.case'#  
                standardized = T)
}

insect_FL <- function(mod, phenotype, covariates = NULL) {
  fitness.landscape(mod, phenotype, covariates, parallel = 'multicore', ncpus = 32,
                    PI.method = 'boot.para',
                    refit.smooth = F,
                    PI.interval = c(0.025,0.975),
                    plt.density = 100)
}

get_insect_FL.df <- function(FL){
  data.frame(mean_fitness = FL$Wbar, FL$points, t(FL$WbarPI)) %>%
    mutate(lower_2.5 = X2.5., upper_97.5 = X97.5.)
}

overdisp_fun <- function(model) {
  ## number of variance parameters in 
  ##   an n-by-n variance-covariance matrix
  vpars <- function(m) {
    nrow(m)*(nrow(m)+1)/2
  }
  model.df <- sum(sapply(VarCorr(model),vpars))+length(fixef(model))
  rdf <- nrow(model.frame(model))-model.df
  rp <- residuals(model,type="pearson")
  Pearson.chisq <- sum(rp^2)
  prat <- Pearson.chisq/rdf
  pval <- pchisq(Pearson.chisq, df=rdf, lower.tail=FALSE)
  c(chisq=Pearson.chisq,ratio=prat,rdf=rdf,p=pval)
}

inverse_logit <- function(x) exp(x)/(1+exp(x))

#Beta_avg_grad <- function(logistic_model, mean_fitness) {
#  brackets <- predict(logistic_model, type = "response") * (1 - predict(logistic_model, type="response"))
#  gradient <- mean(brackets) * fixef(logistic_model) / mean_fitness 
#  return(gradient)
#}

## Calculate directional selection gradients
# following method of Janzen and Stern 1998, Evolution
Beta_avg_grad <- function(.) {
  mean_fitness <- inverse_logit(fixef(update(., .~. -sc.gall_size -sc.clutch_size -sc.female_preference)))
  brackets <- predict(., type = "response") * (1 - predict(., type="response"))
  mean(brackets) * fixef(.)[-1] / mean_fitness 
}

## Bootstrap fitness estimates
bootstrap_fitness <- function(logistic_model, newdata, bootstraps, intervals=c(0.025,0.975)){
  model_matrix <- model.matrix(~ sc.gall_size + sc.clutch_size + sc.female_preference, newdata)

  # Absolute fitness
  get_absolute_fitness <- function(.) inverse_logit(model_matrix %*% fixef(.))

  if(is.null(bootstraps) == FALSE){
    get_bootstraps_absolute_fitness <-bootMer(logistic_model, FUN = get_absolute_fitness, nsim=bootstraps, parallel="multicore", ncpus=32)

   get_intervals_absolute_fitness <- apply(get_bootstraps_absolute_fitness$t, 2, function(x) x[order(x)][c(round(bootstraps*intervals[1],0),round(bootstraps*intervals[2],0))])

  absolute_fitness_df <- data.frame(newdata, 
                    average = get_absolute_fitness(logistic_model), 
                    lower = get_intervals_absolute_fitness[1, ],
                    upper = get_intervals_absolute_fitness[2, ],
                    t(get_bootstraps_absolute_fitness$t))
  } else {
    absolute_fitness_df <- data.frame(newdata, 
                    average = get_absolute_fitness(logistic_model))
  }

  # Relative fitness
  get_relative_fitness <- function(.) get_absolute_fitness(.)/mean(get_absolute_fitness(.))

  if(is.null(bootstraps) == FALSE){
    get_bootstraps_relative_fitness <-bootMer(logistic_model, FUN = get_relative_fitness, nsim=bootstraps, parallel="multicore", ncpus=32)

  get_intervals_relative_fitness <- apply(get_bootstraps_relative_fitness$t, 2, function(x) x[order(x)][c(round(bootstraps*intervals[1],0),round(bootstraps*intervals[2],0))])

  relative_fitness_df <- data.frame(newdata, 
                    average = get_relative_fitness(logistic_model), 
                    lower = get_intervals_relative_fitness[1, ],
                    upper = get_intervals_relative_fitness[2, ],
                    t(get_bootstraps_relative_fitness$t))
  } else {
      relative_fitness_df <- data.frame(newdata, 
                    average = get_relative_fitness(logistic_model))
  }

  # Organize data
  return(list(absolute_fitness = absolute_fitness_df, relative_fitness = relative_fitness_df))
}
```

@@

To characterize the shape of the fitness landscape, we quantified selection gradients acting on each trait in simple vs. complex food webs. We did this by fitting separate statistical models to data from each food-web treatment. We used generalized linear mixed models (GLMMs, @Bolker2009) with larval survival (0 or 1) as our response variable and measure of fitness. We specified linear and quadratic terms for each gall trait as well as linear interaction terms between each gall trait, as fixed effects in the statistical models. To account for the correlated structure of clutch size (gall level) and female preference (plant level) as well as any other independent effects of willow genotype on larval survival, we specified gall ID nested within plant ID nested within plant genotype as random intercepts in our statistical models. Since we were interested in characterizing the fitness landscape -- the relationship between mean trait values and population mean fitness -- we assumed the mean value of our random effects (i.e. setting them to zero) to estimate selection gradients. We then used the method of @Janzen1998 to calculate directional ($\beta_{z_i}$), nonlinear ($\gamma_{z_i}$), and correlational ($\gamma_{z_i,z_j}$) selection gradients and used parametric bootstrapping (1000 replicates) to calculate their 95% confidence intervals (@Bolker2009). To test whether selection gradients differed between treatments, we used our bootstrapped estimates to calculate the probability that selection gradients in the simple food web were larger/smaller than in the complex food web (i.e. the p-value). All analyses and visualizations were conducted in R.

```{r Selection Gradients - Complex Food Web, cache=TRUE}
# note that I've switched to using the lme4 package to analyzing the GLMM

# GLMM ----
glmer_control <- glmer(gall_survival ~ sc.gall_size + sc.clutch_size + sc.female_preference +
                         I(sc.gall_size^2) + I(sc.clutch_size^2) + I(sc.female_preference^2) +
                         sc.gall_size:sc.clutch_size + sc.gall_size:sc.female_preference + sc.clutch_size:sc.female_preference +
                           (1|Genotype/Plant_Position/Gall_Number),
                         data = control_df,
                         family = binomial(link = logit), optimizer = "bobyqa")
# summary(glmer_control)

# Mean fitness ----
Wbar_control <- inverse_logit(fixef(update(glmer_control, .~. -sc.gall_size -sc.clutch_size -sc.female_preference)))

# Confidence intervals for selection gradients ----
boot_Beta_control <- bootMer(glmer_control, FUN = Beta_avg_grad, nsim=100, parallel="multicore", ncpus=32)

## Summary ----
summary_Beta_glmm_control <- data.frame(
  Phenotype = c("Gall size", "Clutch size", "Female preference"),
  Food_Web = "Complex",
  Model = "GLMM",
  Beta = round(Beta_avg_grad(glmer_control),3),
  lower_2.5 = round(apply(boot_Beta_control$t, 2, FUN = function(bootstraps) quantile(bootstraps, 0.025)),3),
  upper_97.5 = round(apply(boot_Beta_control$t, 2, FUN = function(bootstraps) quantile(bootstraps, 0.975)),3))

summary_Beta_glmm_control
```

```{r Selection Gradients - Simple Food Web, cache=TRUE}

## GLMM ----
glmer_treatment <- glmer(gall_survival ~ sc.gall_size + sc.clutch_size + sc.female_preference +
                         I(sc.gall_size^2) + I(sc.clutch_size^2) + I(sc.female_preference^2) +
                         sc.gall_size:sc.clutch_size + 
                          sc.gall_size:sc.female_preference + 
                           sc.clutch_size:sc.female_preference +
                           (1|Genotype/Plant_Position/Gall_Number),
                         data = treatment_df,
                         family = binomial(link = logit), optimizer = "bobyqa")
summary(glmer_treatment)

## Mean fitness ----
Wbar_treatment <- inverse_logit(fixef(update(glmer_treatment, .~. -sc.gall_size -sc.clutch_size -sc.female_preference)))

## Confidence intervals for selection gradients ----
boot_Beta_treatment <- bootMer(glmer_treatment, FUN = Beta_avg_grad, nsim=100, parallel="multicore", ncpus=32)

## Summary ----
summary_Beta_glmm_treatment <- data.frame(
  Phenotype = c("Gall size", "Clutch size", "Female preference"),
  Food_Web = "Simple",
  Model = "GLMM",
  Beta = round(Beta_avg_grad(glmer_treatment),2),
  lower_2.5 = round(apply(boot_Beta_treatment$t, 2, FUN = function(bootstraps) quantile(bootstraps, 0.025)),2),
  upper_97.5 = round(apply(boot_Beta_treatment$t, 2, FUN = function(bootstraps) quantile(bootstraps, 0.975)),2))

summary_Beta_glmm_treatment
```
